\textbf{SPECFEM3D}\\ \textbf{User Manual}

### 2010-11-09 doc/USER_MANUAL/manual_SPECFEM3D.tex
updated the SPECFEM3D manual

The SPECFEM3D software package relies on the Scotch library to partition the meshes created with CUBIT.
Note that we use CUBIT to create meshes of hexahedra, but other packages can be used as well, for instance GiD from http://gid.cimne.upc.es or Gmsh from http://geuz.org/gmsh. Even mesh creation packages that generate tetrahedra, for instance TetGen from http://tetgen.berlios.de, can be used because each tetrahedron can then easily be decomposed into four hexahedra as shown in the picture of the TetGen logo at http://tetgen.berlios.de/figs/Delaunay-Voronoi-3D.gif; while this approach does not generate hexahedra of optimal quality, it can ease mesh creation in some situations and it has been shown that the spectral-element method can very accurately handle distorted mesh elements \citep{OlSe11}.

The path to the SCOTCH library needs to be set with the option \texttt{-{}-with-scotch-lib}. Just as an example:
\begin{lyxcode}
./configure~FC=ifort~MPIFC=mpif90~-{}-with-scotch-lib=/opt/scotch/gcc64/lib
\end{lyxcode}

@@

\begin{description}
\item [{\texttt{F90}}] Path to the Fortran90 compiler.
\item [{\texttt{MPIF90}}] Path to MPI Fortran90.
\item [{\texttt{MPI\_FLAGS}}] Some systems require this flag to link to
MPI libraries.
\item [{\texttt{FLAGS\_CHECK}}] Compiler flag for non-critical subroutines.
\item [{\texttt{FLAGS\_NO\_CHECK}}] Compiler flag for creating fast, production-run
code for critical subroutines.
\end{description}
The \texttt{Makefile} contains a number of suggested entries for various
compilers, e.g., Portland, Intel, Absoft, NAG, and Lahey. The software
has run on a wide variety of compute platforms, e.g., various PC clusters
and machines from Sun, SGI, IBM, Compaq, and NEC. Select the compiler
you wish to use on your system and choose the related optimization
flags. Note that the default flags in the \texttt{Makefile} are undoubtedly
not optimal for your system, so we encourage you to experiment with
these flags and to solicit advice from your systems administrator.
Selecting the right compiler and optimization flags can make a tremendous
difference in terms of performance. We welcome feedback on your experience
with various compilers and flags.

@@

Finally, before compiling make sure that the subdirectories \texttt{obj}
and \texttt{OUTPUT\_FILES} exist within the directory with the source
code (\texttt{SPECFEM3D\_V2.0.0}). The \texttt{go\_mesher} script
discussed in Chapter~\ref{cha:Running-the-Mesher-Meshfem3D} automatically
takes care of creating the \texttt{OUTPUT\_FILES} directory.

@@

\section{\label{cha:Running-the-Mesher-CUBIT}Meshing with \texttt{CUBIT}}
\subsection{Exporting the Mesh}
\subsection{Partitioning the Mesh with \texttt{xdecompose\_mesh\_SCOTCH}}

@@

An efficient partitioning is important because it leverages the overall running time of the application.
It amounts to balance the number of elements in each slice while minimizing the communication coasts
resulting from the placement of adjacent elements to different processors. \texttt{decompose\_mesh\_SCOTCH} depends on the Scotch library
which provides efficent static mapping, graph and mesh partitioning routines. Scotch is a free software package developed by
Fran\c {c}ois Pellegrini et al. from LaBRI and INRIA in Bordeaux, France, downloadable from the web page \texttt{https://gforge.inria.fr/projects/scotch/}. It is more recent than METIS, actively maintained and performs better in many cases.
\\

Prior to compiling \texttt{decompose\_mesh\_SCOTCH}, make sure you have correctly specified the path
of the Scotch library with the option \texttt{-{}-with-scotch-lib} of the \texttt{configure} script.
For convenience we provide the source code of SCOTCH, which is released open source under the French CeCILL-C version 1 license,
in directory src/decompose\_mesh\_SCOTCH/scotch\_5.1.10b. Please refer to file INSTALL.txt in that directory to see how to compile it.
In the future you should be able to find for recent versions at http://www.labri.fr/perso/pelegrin/scotch/scotch\_en.html .
\\

Then you are ready to compile,
in the directory \texttt{decompose\_mesh\_SCOTCH} type '\texttt{make decompose\_mesh\_SCOTCH}'.
If all paths and flags have been set correctly, the executable \texttt{xdecompose\_mesh\_SCOTCH} should be produced.

The partitioning is done in serial for now (in the next release we will provide a parallel version of that code),
the synopsis is:
\begin{lyxcode}
./decompose\_mesh\_SCOTCH  nparts  input\_directory output\_directory
\end{lyxcode}

@@

\noindent \begin{flushleft}
%
\begin{figure}[t]
\noindent \begin{centering}
\includegraphics[scale=0.5]{figures/socal_map_mpi}
\par\end{centering}

@@

You are now ready to compile the internal mesher. This is an alternative to CUBIT
for the mesh generation of relatively simple geological models. The mesher is no longer
dedicated to Southern California and more flexiblity is provided in this version of the package.

In the directory \texttt{meshfem3D} type `\texttt{make meshfem3D}'. If all paths and flags
have been set correctly, the mesher should now compile and produce
the executable \texttt{xmeshfem3D}.

Input for the mesh generation program is provided through the parameter
file \texttt{Par\_file}, which resides in the subdirectory \texttt{meshfem3D/DATA}.
Before running the mesher, a number of parameters need to be set in
the \texttt{Par\_file}. This requires a basic understanding of how
the SEM is implemented, and we encourage you to read \citet{KoVi98,KoTr99}
and \citet{KoLiTrSuStSh04}.

The mesher and the solver use UTM coordinates internally, therefore
you need to define the zone number for the UTM projection (e.g., zone
11 for Los Angeles). Use decimal values for latitude and longitude
(no minutes/seconds). These values are approximate; the mesher will
round them off to define a square mesh in UTM coordinates. When running
benchmarks on rectangular models, turn the UTM projection off by using
the flag \texttt{\small SUPPRESS\_UTM\_PROJECTION}, in which case
all `longitude' parameters simply refer to the $x$~axis, and all
`latitude' parameters simply refer to the $y$~axis. To run the mesher
for a global simulation, the following parameters need to be set in
the \texttt{Par\_file}:

@@

\texttt{material\_ID}  \texttt{rho}  \texttt{vp}  \texttt{vs}  \texttt{Q\_flag}  \texttt{anisotropy\_flag} \texttt{domain\_ID}

@@

     \item \texttt{Q\_flag}           : 0=no attenuation / standard \texttt{Q} attenutation value

@@

Mesher output is provided in the \texttt{OUTPUT\_FILES} directory
in \texttt{output\_mesher.txt}; this file provides lots of details
about the mesh that was generated. Alternatively, output can be directed
to the screen instead by uncommenting a line in \texttt{constants.h}:

@@

The quality of the mesh may be inspected more precisely based upon the serial code
\texttt{check\_mesh\_quality\_CUBIT\_Abaqus.f90}. In the directory \texttt{check\_mesh\_quality\_CUBIT\_Abaqus}, type

@@

Either you previously used  \texttt{xmeshfem3D} or \texttt{xdecompose\_mesh\_SCOTCH}, the next step in the workflow is to compile \texttt{xgenerate\_databases}. This program is going to create all the missing information needed by the SEM solver. In the main directory type '\texttt{make generate\_databases}'. Input for the program is provided through the main parameter file \texttt{Par\_file}, which resides in the subdirectory \texttt{DATA}. Before running \texttt{xgenerate\_databases}, a number of parameters need to be set in the \texttt{Par\_file}:

\begin{description}
\item [{\texttt{SIMULATION\_TYPE}}] is set to 1 for forward simulations,
2 for adjoint simulations (see Section \ref{sec:Adjoint-simulation-finite})
and 3 for kernel simulations (see Section \ref{sec:Finite-Frequency-Kernels}).
\item [{\texttt{SAVE\_FORWARD}}] is only set to \texttt{.true.} for a forward
simulation with the last frame of the simulation saved, as part of
the finite-frequency kernel calculations (see Section \ref{sec:Finite-Frequency-Kernels}).
For a regular forward simulation, leave \texttt{SIMULATION\_TYPE}
and \texttt{SAVE\_FORWARD} at their default values.
\item [{\texttt{\noun{UTM\_PROJECTION\_ZONE}}}] UTM projection zone in
which your model resides, only valid when \texttt{SUPPRESS\_UTM\_}~\\
\texttt{PROJECTION} is \texttt{.false.}.
\item [{\texttt{SUPPRESS\_UTM\_PROJECTION}}] set to be \texttt{.false.}
when your model range is specified in the geographical coordinates,
and needs to be \texttt{.true.} when your model is specified in a
cartesian coordinates. \noun{UTM projection zone in which your simulation
region resides.}
\item [{\texttt{NPROC}}] The number of MPI processors, each one is assigned one slice of the whole mesh.
\item [{\texttt{NSTEP}}] The number of time steps of the simulation. This controls the
length of the numerical simulation, i.e., twice the number of time steps
requires twice as much CPU time. This feature is not used at the time
of meshing but is required for the solver, i.e., you may change this
parameter after running  \texttt{xgenerate\_databases}.
\item [{\texttt{DT}}] The length of each time step in seconds.
If you used xmeshfem3D to generate your mesh, you should set the value suggested in \texttt{meshfem3D/OUTPUT\_FILES/output\_mesher.txt}.

@@

parameter after running xgenerate\_databases.

@@

xgenerate\_databases generates the necessary databases in parallel, one set

@@

files generated by xgenerate\_databases.

@@

the file system (\texttt{timestamp{*}} files in the \texttt{OUTPUT\_FILES}

@@

information about the source time function in the file \texttt{OUTPUT\_FILES/plot\_source\_time\_function.txt}.

@@

Now that you have successfully generated the databases, you are ready to compile
the solver. For reasons of speed, the solver uses static memory allocation.
Therefore it needs to be recompiled (type `\texttt{make clean}' and
`\texttt{make specfem3D}') every time one reruns \texttt{xgenerate\_databases}. To compile
the solver one needs a file created by \texttt{xgenerate\_databases} in the directory
\texttt{OUTPUT\_FILES} called \texttt{values\_from\_mesher.h}, which
contains parameters describing the static size of the arrays as well
as the setting of certain flags.

The solver needs three input files in the \texttt{DATA} directory
to run: the \texttt{Par\_file} which was discussed in detail in Chapters~\ref{cha:Running-the-Mesher-CUBIT} and~\ref{cha:Running-the-Mesher-Meshfem3D},
the earthquake source parameter file \texttt{CMTSOLUTION}, and the
stations file \texttt{STATIONS}. Most parameters in the \texttt{Par\_file}
should be set prior to running the mesher. Only the following parameters
may be changed after running \texttt{xgenerate\_databases}:

\begin{itemize}
\item the simulation type control parameters: \texttt{SIMULATION\_TYPE}
and \texttt{SAVE\_FORWARD}
\item the time step parameters \texttt{NSTEP} and \texttt{DT}
\item the movie control parameters \texttt{MOVIE\_SURFACE}, \texttt{MOVIE\_VOLUME},
and \texttt{NTSTEPS\_BETWEEN\_FRAMES}
\item the ShakeMap\textregistered option \texttt{CREATE\_SHAKEMAP}
\item the output information parameters \texttt{NTSTEP\_BETWEEN\_OUTPUT\_INFO}
and \texttt{NTSTEP\_BETWEEN\_OUTPUT\_}~\\
\texttt{SEISMOS}
\item the \texttt{PRINT\_SOURCE\_TIME\_FUNCTION} flags
\end{itemize}
Any other change to the \texttt{Par\_file} implies rerunning both
the mesher and the solver.

@@

\begin{lyxcode}
{\small Station~Network~Latitude~(degrees)~Longitude~(degrees)~Elevation~(m)~burial~(m)~}{\small \par}
\end{lyxcode}
The mesher filters the list of stations in file \texttt{DATA/STATIONS}
to exclude stations that are not located within the region given in
the \texttt{Par\_file} (between \texttt{LATITUDE\_MIN} and \texttt{LATITUDE\_MAX}
and between \texttt{LONGITUDE\_MIN} and \texttt{LONGITUDE\_MAX}).
The filtered file is called \texttt{DATA/STATIONS\_FILTERED}.

Solver output is provided in the \texttt{OUTPUT\_FILES} directory
in the \texttt{output\_solver.txt} file. Output can be directed to
the screen instead by uncommenting a line in \texttt{constants.h}:

@@

While the solver is running, its progress may be tracked by monitoring
the `\texttt{\small timestamp{*}}' files in the \texttt{\small OUTPUT\_FILES}
directory. These tiny files look something like this:

@@

\begin{lyxcode}
UTILS/change\_simulation\_type.pl~-f~
\end{lyxcode}
and then collect the recorded seismograms at all the stations given
in \texttt{DATA/STATIONS}.

>>>>

The SPECFEM3D software package relies on the Scotch library to partition the meshes created with CUBIT.
Note that we use CUBIT to create meshes of hexahedra, but other packages can be used as well, for instance GiD from \url{http://gid.cimne.upc.es} or Gmsh from \url{http://geuz.org/gmsh}. Even mesh creation packages that generate tetrahedra, for instance TetGen from \url{http://tetgen.berlios.de}, can be used because each tetrahedron can then easily be decomposed into four hexahedra as shown in the picture of the TetGen logo at \url{http://tetgen.berlios.de/figs/Delaunay-Voronoi-3D.gif}; while this approach does not generate hexahedra of optimal quality, it can ease mesh creation in some situations and it has been shown that the spectral-element method can very accurately handle distorted mesh elements \citep{OlSe11}.

The path to the SCOTCH installation needs to be set with the option \texttt{-{}-with-scotch-path}. Just as an example:
\begin{lyxcode}
./configure~FC=ifort~MPIFC=mpif90~-{}-with-scotch-path=/opt/scotch
\end{lyxcode}

@@

\begin{description}
\item [{\texttt{F90}}] Path to the Fortran90 compiler.
\item [{\texttt{MPIF90}}] Path to MPI Fortran90.
\item [{\texttt{MPI\_FLAGS}}] Some systems require this flag to link to
MPI libraries.
\item [{\texttt{FLAGS\_CHECK}}] Compiler flag for non-critical subroutines.
\item [{\texttt{FLAGS\_NO\_CHECK}}] Compiler flag for creating fast, production-run
code for critical subroutines.
\end{description}
The configuration scripts automatically creates for each executable a corresponding
\texttt{Makefile} in the \texttt{src/} subdirectoy.
The \texttt{Makefile} contains a number of suggested entries for various
compilers, e.g., Portland, Intel, Absoft, NAG, and Lahey. The software
has run on a wide variety of compute platforms, e.g., various PC clusters
and machines from Sun, SGI, IBM, Compaq, and NEC. Select the compiler
you wish to use on your system and choose the related optimization
flags. Note that the default flags in the \texttt{Makefile} are undoubtedly
not optimal for your system, so we encourage you to experiment with
these flags and to solicit advice from your systems administrator.
Selecting the right compiler and optimization flags can make a tremendous
difference in terms of performance. We welcome feedback on your experience
with various compilers and flags.

@@

% obsolete
%Finally, before compiling make sure that the subdirectories \texttt{obj}
%and \texttt{OUTPUT\_FILES} exist within the directory with the source
%code (\texttt{SPECFEM3D\_V2.0.0}). The \texttt{go\_mesher} script
%discussed in Chapter~\ref{cha:Running-the-Mesher-Meshfem3D} automatically
%takes care of creating the \texttt{OUTPUT\_FILES} directory.

@@

The first step in running a spectral-element method computation consists of constructing a high-quality mesh for the region under consideration. We provide two possibilities to do so: (1) relying on an external, hexahedral mesher CUBIT or (2) using the provided, internal mesher \texttt{xmeshfem3D}. In the following, we explain these two approaches.

\section{\label{cha:Running-the-Mesher-CUBIT}Meshing with \texttt{CUBIT}}

CUBIT is a meshing tool suite for the creation of finite-element meshes for arbitrarily shaped models. It has been developed and maintained at Sandia National Laboratories and can be purchased for a small academic institutional fee at \url{http://cubit.sandia.gov}.
Our experience showed that using CUBIT greatly facilitates and speeds up the generation and preparation of hexahedral, conforming meshes for a variety of geophysical models with increasing complexity.

\begin{figure}[htbp]
\noindent \begin{centering}
\includegraphics[width=.6\textwidth]{figures/mount-cubit.jpg}
\par\end{centering}

\caption{Example of the graphical user interface of CUBIT. The hexahedral mesh shown in the main display consists of a hexahedral discretization of a single volume with topography.}

\label{fig:mount.cubit}
\end{figure}

The basic steps in creating a load-balanced, partitioned mesh with CUBIT are: 
\begin{description}
\item [1.] setting up a hexahedral mesh with CUBIT, 
\item [2.] exporting the CUBIT mesh into a specfem3d file format and 
\item [3.] partitioning the specfem3d mesh files for a chosen number of cores. 
\end{description}

Examples are provided in the SPECFEM3D package in the subdirectory \texttt{examples/}. Please feel free to contribute your own example to this package by contacting the CIG Computational Seismology Mailing List  \url{cig-seismo@geodynamics.org}.

\subsection{Creating the Mesh with CUBIT}

For the installation and handling of the CUBIT meshing tool suite, please refer to the CUBIT user manual and documentation. 
In order to give you a basic understanding of how to use CUBIT for our purposes, examples are provided in the SPECFEM3D package in the subdirectory \texttt{examples/}. For your convenience, the following examples are provided in the subdirectory \texttt{examples/}:
\begin{description}
\item [{\texttt{homogeneous\_halfspace}}] Basic example creates a single block model and assigns elastic material parameters.
\item [{\texttt{layered\_halfspace}}]  Combines two different, elastic material volumes and creates a refinement layer between the two. This example can be compared for validation against the solutions provided in subdirectory ~\\
\texttt{VALIDATION\_3D\_SEM\_SIMPLER\_LAYER\_SOURCE\_DEPTH/}.
\item [{\texttt{waterlayered\_halfspace}}] Combines an acoustic and elastic material volume as in a schematic marine survey example.
\item [{\texttt{tomographic\_model}}] Creates a single block model whose material properties will have to be read in from a tomographic model file during the databases creation by \texttt{xgenerate\_databases}. 
\end{description}

\begin{figure}[htbp]
\noindent \begin{centering}
\includegraphics[width=.45\textwidth]{figures/example-homogeneous.jpg}
\includegraphics[width=.45\textwidth]{figures/example-2layers.jpg} \\
\includegraphics[width=.45\textwidth]{figures/example-water.jpg}
\includegraphics[width=.45\textwidth]{figures/example-tomo.jpg}
\par\end{centering}

\caption{Screenshots of the CUBIT examples provided in subdirectory  \texttt{examples/}: homogeneous halfspace, layered halfspace, waterlayered halfspace and tomographic model.}

\label{fig:examples.cubit}
\end{figure}

In each example subdirectory you will find a \texttt{README} file, which explains in a step-by-step tutorial the workflow for the example.
Please feel free to contribute your own example to this package by contacting the CIG Computational Seismology Mailing List  \url{cig-seismo@geodynamics.org}.

\subsection{Exporting the Mesh with \texttt{cubit2specfem3d.py} }

Once the geometric model volumes in CUBIT are meshed, you prepare the model for the exportation with the definition of the material blocks and of the boundary surfaces. Thus, prior to exporting the mesh, you need to define blocks specifying the materials and absorbing boundaries in CUBIT.
This process could be done automatically using the script \texttt{boundary\_definition.py} if the mesh meets some conditions or manually, following the block convention:

\begin{description}
\item [material\_name] Each material should have a specific block defined by a unique name. The name convention of the material is to start with either 'elastic' or 'acoustic'. It must be then followed by a unique identifier, e.g. 'elastic 1', 'elastic 2', etc.
The additional attributes to the block define the material description.

For an elastic material:
\begin{description}
\item [material\_id] An integer value which is unique for this material. 
\item [Vp] P-Velocity of the material (given in m/s).
\item [Vs] S-Velocity of the material (given in m/s).
\item [rho] density of the material (given in kg/$m^3$).
\item [Q] quality factor to use in case of a simulation with attenuation turned on. It should be between 1 and 9000. In case no attenuation information is available, it can be set to zero. Please note, that the Vp- and Vs-velocities are given for a reference frequency. To change this reference frequency, you change the value of 
\texttt{ATTENUATION\_f0\_REFERENCE} in the main constants file \texttt{constants.h} found in subdirectory \texttt{src/shared/}.
\item [anisotropic\_flag] Flag describing the anisotropic model to use in case an anisotropic simulation should be conducted. See the file \texttt{model\_aniso.f90} in subdirectory \texttt{src/generate\_databases/} for an implementation of the anisotropic models. In case no anisotropy is available, it can be set to zero.
\end{description}
Note that this material block has to be defined using all the volumes which belong to this elastic material. For volumes belonging to another, different material, you will need to define a new material block.

For an acoustic material:
\begin{description}
\item [material\_id] An integer value which is unique for this material. 
\item [Vp] P-Velocity of the material (given in m/s).
\item [0] S-Velocity of the material is ignored.
\item [rho] density of the material (given in kg/$m^3$).
\end{description}

\item [face\_topo] Block definition for the surface which defines the free surface (which can have topography). The name of this block must be 'face\_topo', the block has to be defined using all the surfaces which constitute the complete free surface of the model.

\item [face\_abs\_xmin] Block definition for the faces on the absorbing boundaries, one block for each surface with x=Xmin. 
\item [face\_abs\_xmax] Block definition for the faces on the absorbing boundaries, one block for each surface with x=Xmax.
\item [face\_abs\_ymin] Block definition for the faces on the absorbing boundaries, one block for each surface with y=Ymin.
\item [face\_abs\_ymax] Block definition for the faces on the absorbing boundaries, one block for each surface with y=Ymax.
\item [face\_abs\_bottom] Block definition for the faces on the absorbing boundaries, one block for each surface with z=bottom.
\end{description}

\begin{figure}[htbp]
\noindent \begin{centering}
\includegraphics[width=.45\textwidth]{figures/mount-surface.jpg}
\includegraphics[width=.45\textwidth]{figures/mount-abs.jpg}
\par\end{centering}

\caption{Example of the block definitions for the free surface 'face\_topo' (left) and the absorbing boundaries, defined in a single block 'face\_abs\_xmin' (right) in CUBIT.}

\label{fig:mount.abs}
\end{figure}

Optionally, instead of specifying for each surface at the boundaries a single block like mentioned above, you can also specify a single block for all boundary surfaces and name it as one of the absorbing blocks above, e.g. 'face\_abs\_xmin'.

After the block definitions are done, you export the mesh using the script \texttt{cubit2specfem3d.py} provided in each of the example directories. If the export was successful, you should find the following files a the subdirectory \texttt{MESH/}:
\begin{description}
\item [nummaterial\_velocity\_file] Defines the material properties. For fully defined materials, the formats is:
\begin{lyxcode}
\texttt{domain\_ID} \texttt{material\_ID}  \texttt{rho}  \texttt{vp}  \texttt{vs}  \texttt{Q}  \texttt{anisotropy\_flag} 
\end{lyxcode}
where 
\texttt{domain\_ID} is 1 for acoustic or 2 for elastic materials, \texttt{material\_ID} a unique identifier,  \texttt{rho} the density in $kg \, m^{-3}$,  \texttt{vp} the P-velocity in $m \, s^{-1}$, \texttt{Vs}  the S-velocity in $m \, s^{-1}$, \texttt{Q} the quality factor and \texttt{anisotropy\_flag} an identifier for anisotropic models.
For a tomographic model, the material definition format is: 
\begin{lyxcode}
\texttt{domain\_ID} \texttt{material\_ID}  \texttt{tomography}  \texttt{name}  
\end{lyxcode}
where
\texttt{domain\_ID} is 1 for acoustic or 2 for elastic materials, \texttt{material\_ID} a negative, unique identifier, \texttt{tomography} keyword for tomographic material definition and \texttt{name} the name of the tomography file. The name is not used so far, rather change the filename defined in the file ~\\
\texttt{src/generate\_databases/model\_tomography.f90}.
\item [materials\_file] Contains the material associations for each element. 
\item [nodes\_coords\_file] Contains the point locations in Cartesian coordinates of the mesh element corners.
\item [mesh\_file] Contains the mesh element connectivity.
\item [free\_surface\_file] Contains the free surface connectivity.
\item [absorbing\_surface\_file\_xmax] Contains the surface connectivity of the absorbing boundary surface at the Xmax.
\item [absorbing\_surface\_file\_xmin] Contains the surface connectivity of the absorbing boundary surface at the Xmin.
\item [absorbing\_surface\_file\_ymax] Contains the surface connectivity of the absorbing boundary surface at the Ymax.
\item [absorbing\_surface\_file\_ymin] Contains the surface connectivity of the absorbing boundary surface at the Ymin.
\item [absorbing\_surface\_file\_bottom] Contains the surface connectivity of the absorbing boundary surface at the bottom.
\end{description}

\subsection{Partitioning the Mesh with \texttt{xdecompose\_mesh\_SCOTCH}}

@@

An efficient partitioning is important because it leverages the overall running time of the application.
It amounts to balance the number of elements in each slice while minimizing the communication coasts
resulting from the placement of adjacent elements to different processors. \texttt{decompose\_mesh\_SCOTCH} depends on the Scotch library
which provides efficent static mapping, graph and mesh partitioning routines. Scotch is a free software package developed by
Fran\c {c}ois Pellegrini et al. from LaBRI and INRIA in Bordeaux, France, downloadable from the web page \url{https://gforge.inria.fr/projects/scotch/}. It is more recent than METIS, actively maintained and performs better in many cases.
\\

Prior to compiling \texttt{decompose\_mesh\_SCOTCH}, make sure you have correctly specified the path
of the Scotch library with the option \texttt{-{}-with-scotch-path} of the \texttt{configure} script.
For convenience we provide the source code of SCOTCH, which is released open source under the French CeCILL-C version 1 license,
in directory ~\\
\texttt{src/decompose\_mesh\_SCOTCH/scotch\_5.1.10b}. Please refer to file INSTALL.txt in that directory to see how to compile it.
In the future you should be able to find for recent versions at  \url{http://www.labri.fr/perso/pelegrin/scotch/scotch\_en.html} .
\\

\begin{figure}[htbp]
\noindent \begin{centering}
\includegraphics[width=.45\textwidth]{figures/mount-partitions.jpg}
\includegraphics[width=.45\textwidth]{figures/mount-partitions2.jpg}
\par\end{centering}

\caption{Example of a mesh partitioning onto four cores. Each single core partition is colored differently. The executable \texttt{xdecompose\_mesh\_SCOTCH} can equally distribute the mesh on any arbitrary number of cores.}

\label{fig:mount.partitions}
\end{figure}

Then you are ready to compile,
in the main directory type '\texttt{make decompose\_mesh\_SCOTCH}'.
If all paths and flags have been set correctly, the executable \texttt{bin/xdecompose\_mesh\_SCOTCH} should be produced.

The partitioning is done in serial for now (in the next release we will provide a parallel version of that code),
the synopsis is:
\begin{lyxcode}
./bin/xdecompose\_mesh\_SCOTCH  nparts  input\_directory output\_directory
\end{lyxcode}

@@

\noindent \begin{flushleft}
%
\begin{figure}[htbp]
\noindent \begin{centering}
\includegraphics[scale=0.5]{figures/socal_map_mpi}
\par\end{centering}

@@

In case you successfully ran the configuration script, you are also ready to compile the internal mesher. 
This is an alternative to CUBIT
for the mesh generation of relatively simple geological models. The mesher is no longer
dedicated to Southern California and more flexiblity is provided in this version of the package.

In the main directory type `\texttt{make meshfem3D}'. If all paths and flags
have been set correctly, the mesher should now compile and produce
the executable \texttt{bin/xmeshfem3D}.

Input for the mesh generation program is provided through the parameter
file \texttt{Mesh\_Par\_file}, which resides in the subdirectory \texttt{in\_data\_files/meshfem3D\_files/}.
Before running the mesher, a number of parameters need to be set in
the \texttt{Mesh\_Par\_file}. This requires a basic understanding of how
the SEM is implemented, and we encourage you to read \citet{KoVi98,KoTr99}
and \citet{KoLiTrSuStSh04}.

The mesher and the solver use UTM coordinates internally, therefore
you need to define the zone number for the UTM projection (e.g., zone
11 for Los Angeles). Use decimal values for latitude and longitude
(no minutes/seconds). These values are approximate; the mesher will
round them off to define a square mesh in UTM coordinates. When running
benchmarks on rectangular models, turn the UTM projection off by using
the flag \texttt{\small SUPPRESS\_UTM\_PROJECTION}, in which case
all `longitude' parameters simply refer to the $x$~axis, and all
`latitude' parameters simply refer to the $y$~axis. To run the mesher
for a global simulation, the following parameters need to be set in
the \texttt{Mesh\_Par\_file}:

@@

\texttt{material\_ID}  \texttt{rho}  \texttt{vp}  \texttt{vs}  \texttt{Q}  \texttt{anisotropy\_flag} \texttt{domain\_ID}

@@

     \item \texttt{Q}           : 0=no attenuation / standard \texttt{Q} attenutation value

@@

Mesher output is provided in the \texttt{in\_out\_files/OUTPUT\_FILES} directory
in \texttt{output\_mesher.txt}; this file provides lots of details
about the mesh that was generated. Alternatively, output can be directed
to the screen instead by uncommenting a line in \texttt{constants.h}:

@@

The quality of the mesh may be inspected more precisely based upon the serial code
in the file  ~\\
\texttt{check\_mesh\_quality\_CUBIT\_Abaqus.f90}. In the directory \texttt{check\_mesh\_quality\_CUBIT\_Abaqus}, type

@@

Either you previously used  \texttt{xmeshfem3D} or \texttt{xdecompose\_mesh\_SCOTCH}, the next step in the workflow is to compile \texttt{xgenerate\_databases}. This program is going to create all the missing information needed by the SEM solver.

\begin{figure}[htbp]
\noindent \begin{centering}
\includegraphics[width=.6\textwidth]{figures/workflow.jpg}
\par\end{centering}

\caption{Schematic workflow for a SPECFEM3D simulation. The exectable \texttt{xgenerate\_databases} creates the GLL mesh points and assigns the specific model parameters.}

\label{fig:workflow.databases}
\end{figure}

In the main directory type '\texttt{make generate\_databases}'. Input for the program is provided through the main parameter file \texttt{Par\_file}, which resides in the subdirectory \texttt{in\_data\_files/}. Before running \texttt{xgenerate\_databases}, a number of parameters need to be set in the \texttt{Par\_file}:

\begin{description}
\item [{\texttt{SIMULATION\_TYPE}}] is set to 1 for forward simulations,
2 for adjoint simulations (see Section \ref{sec:Adjoint-simulation-finite})
and 3 for kernel simulations (see Section \ref{sec:Finite-Frequency-Kernels}).
\item [{\texttt{SAVE\_FORWARD}}] is only set to \texttt{.true.} for a forward
simulation with the last frame of the simulation saved, as part of
the finite-frequency kernel calculations (see Section \ref{sec:Finite-Frequency-Kernels}).
For a regular forward simulation, leave \texttt{SIMULATION\_TYPE}
and \texttt{SAVE\_FORWARD} at their default values.
\item [{\texttt{\noun{UTM\_PROJECTION\_ZONE}}}] UTM projection zone in
which your model resides, only valid when \texttt{SUPPRESS\_UTM\_}~\\
\texttt{PROJECTION} is \texttt{.false.}.
\item [{\texttt{SUPPRESS\_UTM\_PROJECTION}}] set to be \texttt{.false.}
when your model range is specified in the geographical coordinates,
and needs to be \texttt{.true.} when your model is specified in a
cartesian coordinates. \noun{UTM projection zone in which your simulation
region resides.}
\item [{\texttt{NPROC}}] The number of MPI processors, each one is assigned one slice of the whole mesh.
\item [{\texttt{NSTEP}}] The number of time steps of the simulation. This controls the
length of the numerical simulation, i.e., twice the number of time steps
requires twice as much CPU time. This feature is not used at the time
of meshing but is required for the solver, i.e., you may change this
parameter after running  \texttt{xgenerate\_databases}.
\item [{\texttt{DT}}] The length of each time step in seconds.
If you used xmeshfem3D to generate your mesh, you should set the value suggested in \texttt{in\_out\_files/OUTPUT\_FILES/output\_mesher.txt}.

@@

parameter after running \texttt{xgenerate\_databases}.

@@

\texttt{xgenerate\_databases} generates the necessary databases in parallel, one set

@@

files generated by \texttt{xgenerate\_databases}.

@@

the file system (\texttt{timestamp{*}} files in the \texttt{in\_out\_files/OUTPUT\_FILES}

@@

information about the source time function in the file \texttt{in\_out\_files/OUTPUT\_FILES/plot\_source\_time\_function.txt}.

@@

Now that you have successfully generated the databases, you are ready to compile
the solver. In the main directory type `\texttt{make specfem3D}'.
%obsolete
%For reasons of speed, the solver uses static memory allocation.
%Therefore it needs to be recompiled (type `\texttt{make clean}' and
%`\texttt{make specfem3D}') every time one reruns \texttt{xgenerate\_databases}. To compile
%the solver one needs a file created by \texttt{xgenerate\_databases} in the directory
%\texttt{OUTPUT\_FILES} called \texttt{values\_from\_mesher.h}, which
%contains parameters describing the static size of the arrays as well
%as the setting of certain flags.

The solver needs three input files in the \texttt{in\_data\_files/} directory
to run: 
\begin{description}
\item [{\texttt{Par\_file}}] the main parameter file which was discussed in detail in the previous 
Chapter~\ref{cha:Creating-Distributed-Databases},
\item [{\texttt{CMTSOLUTION}}] the earthquake source parameter file, and 
\item [{\texttt{STATIONS}}] the stations file. 
\end{description}

Most parameters in the \texttt{Par\_file}
should be set prior to running the databases generation. Only the following parameters
may be changed after running \texttt{xgenerate\_databases}:

\begin{itemize}
\item the simulation type control parameters: \texttt{SIMULATION\_TYPE}
and \texttt{SAVE\_FORWARD}
\item the time step parameters \texttt{NSTEP} and \texttt{DT}
\item the movie control parameters \texttt{MOVIE\_SURFACE}, \texttt{MOVIE\_VOLUME},
and \texttt{NTSTEPS\_BETWEEN\_FRAMES}
\item the ShakeMap\textregistered option \texttt{CREATE\_SHAKEMAP}
\item the output information parameters \texttt{NTSTEP\_BETWEEN\_OUTPUT\_INFO}
and \texttt{NTSTEP\_BETWEEN\_OUTPUT\_}~\\
\texttt{SEISMOS}
\item the \texttt{PRINT\_SOURCE\_TIME\_FUNCTION} flags
\end{itemize}
Any other change to the \texttt{Par\_file} implies rerunning both
the generate\_databases and the solver.

@@

\begin{lyxcode}
{\small Station~Network~Latitude~(degrees)~Longitude~(degrees)~Elevation~(m)~burial~(m)~}{\small \par}
\end{lyxcode}
The mesher filters the list of stations in file \texttt{in\_data\_files/STATIONS}
to exclude stations that are not located within the region given in
the \texttt{Par\_file} (between \texttt{LATITUDE\_MIN} and \texttt{LATITUDE\_MAX}
and between \texttt{LONGITUDE\_MIN} and \texttt{LONGITUDE\_MAX}).
The filtered file is called \texttt{in\_data\_files/STATIONS\_FILTERED}.

Solver output is provided in the \texttt{in\_out\_files/OUTPUT\_FILES} directory
in the \texttt{output\_solver.txt} file. Output can be directed to
the screen instead by uncommenting a line in \texttt{constants.h}:

@@

While the solver is running, its progress may be tracked by monitoring
the `\texttt{\small timestamp{*}}' files in the ~\\
\texttt{\small in\_out\_files/OUTPUT\_FILES} directory. 
These tiny files look something like this:

@@

\begin{lyxcode}
UTILS/change\_simulation\_type.pl~-f~
\end{lyxcode}
and then collect the recorded seismograms at all the stations given
in \texttt{in\_data\_files/STATIONS}.

\item Then select the stations for which you want to compute the time-reversed
adjoint sources and run the adjoint simulation, and compile them into
the \texttt{in\_data\_files/STATIONS\_ADJOINT} file, which has the same format
as the regular \texttt{in\_data\_files/STATIONS} file.
